agdarsec -- Total Parser Combinators

### 2017-10-30 doc/paper/main.tex
[ doc ] ship it!

These two definitions are only somewhat satisfactory: the
resut of the \parser{digit} parser is still \emph{stringly-typed}.
Instead of using a predicate to decide whether to keep the
value, we can opt for a validation function of type
\mbox{\textit{a → Maybe b}} which returns
a witness whenever the check succeeds. To define this
alternative \parser{guardM} we can once more rely on code
already part of the standard library.

@@

\subsection{Code Reuse: Recognizing Structures}

We have seen how we can already rely on the standard library
of the host language to seemlessly implement combinators.
We can leverage even more of the existing codebase by noticing
that the type constructor \type{Parser} is a \type{Functor},
an \type{Applicative}~\cite{mcbride2008applicative}, a \type{Monad} and also an \type{Alternative}.

@@

\subsection{Expressivity: Higher Order Parsers and Fixpoints}

@@

\shgrab{some}
\hgrab{many}

\remark{Non-Commutativity} The disjunction combinator is
non-commutative. As such the definitions of \textit{some} and
\textit{many} will try to produce the longest list as possible
as opposed to a flipped version of \text{many} which would start
by returning the empty list and slowly offer longer and longer
matches.

@@

This approach can be generalised when defining more complex
languages by having even more tiers, one for each
\emph{fixity level}, see for instance~\ref{example:expr}.
An extended language of arithmetic expressions would for
instance distinguish the level at which addition and
substraction live from the one at which multiplication and
division do.

Our issue with this solution is twofold. Firt, although we did
eventually managed to build a parser that worked as expected,
the compiler was unable to warn us and guide us towards this
correct solution. Additionnally, the blatant partiality of
some of these definitions means that these combinators and
these types are wholly unsuitable in a total setting. We could, of
course use an escape hatch and implement our parsers in Haskell
but that would both be unsafe and mean we would not be able
to run them at typechecking time which we may want to do if
we embed checked examples in our software's documentation,
or use constant values (e.g. filepaths).

@@

\sagrab{Relation/Unary/Indexed}{arrow}
\agrab{Relation/Unary/Indexed}{product}
\agrab{Relation/Unary/Indexed}{constant}

@@

\agrab{Text/Parser/Combinators}{fmap}\vspace*{-2em}
\agrab{Text/Parser/Combinators}{fail}\vspace*{-2em}
\agrab{Text/Parser/Combinators}{disjunction}

@@

\remark{Non-compositionality}\label{parser:some} The higher-order parser \AF{some}
expects a \emph{fully} defined parser as an argument. This makes
it impossible to use it as one of the building blocks of a larger,
recursive parser. Ideally we would rather have a combinator of
type \mbox{\AF{[} \AF{Parser} \AB{A} \AF{⟶} \AF{Parser} (\AF{List$^+$} \AB{A}) \AF{]}}.
This will be adressed in the next subsection.
\medskip{}

@@

\section{Left and Right Chains}

@@

\subsection{Examples}

\todo{expr}\label{example:expr}

>>>>

These two definitions are only somewhat satisfactory: the
result of the \parser{digit} parser is still \emph{stringly-typed}.
Instead of using a predicate to decide whether to keep the
value, we can opt for a validation function of type
\mbox{\textit{a → Maybe b}} which returns
a witness whenever the check succeeds. To define this
alternative \parser{guardM} we can once more rely on code
already part of the standard library.

@@

\subsection{Expressivity: Structures, Higher Order Parsers and Fixpoints}

%\subsection{Code Reuse: Recognizing Structures}

We have seen how we can already rely on the standard library
of the host language to seamlessly implement combinators.
We can leverage even more of the existing codebase by noticing
that the type constructor \type{Parser} is a \type{Functor},
an \type{Applicative}~\cite{mcbride2008applicative}, a \type{Monad} and also an \type{Alternative}.

@@

%\subsection{Expressivity: Higher Order Parsers and Fixpoints}

@@

\begin{minipage}{0.45\textwidth}
\hgrab{some}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\hgrab{many}
\end{minipage}

\remark{Non-Commutative} The disjunction combinator is
non-commutative. As such the definitions of \textit{some} and
\textit{many} will try to produce the longest list as possible
as opposed to a flipped version of \text{many} which would start
by returning the empty list and slowly offer longer and longer
matches.

@@

This approach can be generalised when defining more complex
languages by having even more tiers, one for each
\emph{fixity level}\label{fixity}, see for instance~\ref{example:expr}.
An extended language of arithmetic expressions would for
instance distinguish the level at which addition and
substraction live from the one at which multiplication and
division do.

Our issue with this solution is twofold. First, although we did
eventually managed to build a parser that worked as expected,
the compiler was unable to warn us and guide us towards this
correct solution. Additionally, the blatant partiality of
some of these definitions means that these combinators and
these types are wholly unsuitable in a total setting. We could, of
course use an escape hatch and implement our parsers in Haskell
but that would both be unsafe and mean we would not be able
to run them at typechecking time which we may want to do if
we embed checked examples in our software's documentation,
or use constant values (e.g. filepaths).

@@

\sagrab{Relation/Unary/Indexed}{arrow}

\begin{minipage}{0.55\textwidth}
  \agrab{Relation/Unary/Indexed}{product}
\end{minipage}
\begin{minipage}{0.30\textwidth}
\agrab{Relation/Unary/Indexed}{constant}
\end{minipage}

@@

\agrab{Text/Parser/Combinators}{fmap}
\begin{minipage}{0.30\textwidth}
  \agrab{Text/Parser/Combinators}{fail}
\end{minipage}
\begin{minipage}{0.55\textwidth}
  \agrab{Text/Parser/Combinators}{disjunction}
\end{minipage}

@@

\remark{Non-Compositional}\label{parser:some} The higher-order parser \AF{some}
expects a \emph{fully} defined parser as an argument. This makes
it impossible to use it as one of the building blocks of a larger,
recursive parser. Ideally we would rather have a combinator of
type \mbox{\AF{[} \AF{Parser} \AB{A} \AF{⟶} \AF{Parser} (\AF{List$^+$} \AB{A}) \AF{]}}.
This will be addressed in the next subsection.
\medskip{}

@@

\section{Left Chains}

@@

\section{Fully Worked-Out Example}

From \AF{hchainl}, one can derive \AF{chainl1} which is not heterogeneous
and uses the same parser for the seed and the constructors' arguments.
This combinator together with the idea of fixity mentioned in Section~\ref{fixity}
is typically used to implement left-recursive grammars. Looking up the
documentation of the \texttt{parsec} library on hackage~\cite{doc:parsec}
we can find a fine example: an extension of our early arithmetic language.

\shgrab{example}

One important thing to note here is that in the end we not only get a
parser for the \type{expr}essions but also each one of the intermediate
categories \type{term} and \type{factor}. Luckily, our library lets us
take fixpoints of any sized types we may fancy. As such, we can define
a sized record of parsers for each one of the syntactic categories:

\agrab{Text/Parser/Examples/Expr}{language}

Here, unlike the Haskell example, we decide to be painfully explicit
about the syntactic categories we are considering: we mutually define
three inductive types representing \emph{left-associated} arithmetic
expressions.

\begin{minipage}{0.33\textwidth}
  \agrab{Text/Parser/Examples/Expr}{expr}
\end{minipage}
\begin{minipage}{0.35\textwidth}
  \agrab{Text/Parser/Examples/Expr}{term}
\end{minipage}
\begin{minipage}{0.20\textwidth}
  \agrab{Text/Parser/Examples/Expr}{factor}
\end{minipage}

The definition of the parser itself is then basically the same as the
Haskell one. Contrary to a somewhat popular belief, working in a
dependently-typed language does not force us to add any type annotation
except for the top-level one.

\agrab{Text/Parser/Examples/Expr}{parser}

We can notice four minor changes with respect to the Haskell version. Firstly,
the intermediate parsers need to be declared before being used which effectively
reverses the order in which they are spelt out. Secondly, the recursive calls
are now explicit: in the definition of \type{factor}, \AF{expr} is \AF{map}ped
under the \BOX{} to project the recursive call to the \AD{Expr} \AR{Parser} out
of \AR{Language}. Thirdly, we use \AF{hchainl} instead of \AF{chainl1} because
breaking the grammar into three distinct categories leads us to parsing
\emph{h}eterogeneous left chains. Fourthly, we have to insert calls to \AF{box}
to lift \AR{Parser}s into boxed ones whenever the added guarantee that the call
will be guarded is of no use to us.

\section{More Power: Switching to other Representations}

We have effectively managed to take Haskell's successful approach to defining
a Domain Specific Language of parser combinators and impose type constraints
which make it safe to use in a total setting. All of which we have done whilst
keeping the concision and expressivity of the original libraries. A natural
next question would be the speed and efficiency of such a library.

Although we have been using a concrete type for \AR{Parser} throughout
this article, our library actually implements a more general one. It uses
Agda's instance arguments throughout thus letting the user pick the
representation they like best.

Firstly, there is nothing special about vectors of characters as an input
type: any sized input off of which one can peel characters one at a time
would do. Users may instead use Haskell's \type{Text} packaged together
with an irrelevant proof that the given text has the right length and a
binding for \textit{uncons}. This should lead to a more efficient memory
representation of the text being analysed.

Similarly, there is no reason to force the user to get back a \emph{list}
of successes: any alternative monad will do. This means in particular that
a user may for instance instrument a parser with a logging ability to be
able to return good error messages, have a (re)configurable grammar using
a \type{Reader} transformer or use a \type{Maybe} type if they want to make
explicit the fact that their grammar is unambiguous.

\section{Related Work}

This work is based on Hutton and Meijer's influential functional pearl~\citeyear{hutton1998monadic}
which builds on Walder's insight that exception handling and backtracking can be
realised using a list of successes~\citeyear{wadler1985replace}. Similar Domain
Specific Languages have been implemented in various functional languages such as
Scala~\cite{moors2008parser} or, perhaps more interestingly for us, Rust~\cite{couprie2015nom}
where the added type-level information about ownership can help implement a
guaranteed zero-copy parser.

\subsection{Total Parser Combinators}

When it comes to total programming languages, Danielsson's library~\citeyear{danielsson2010total}
is to our knowledge the only attempt so far at defining a library of total
parser combinators in a dependently-typed host language. He reifies recursive
grammars as values of a mixed inductive-coinductive type and tracks at the
type level whether a sub-grammar accepts the empty word and, as a consequence,
whether one can meaningfully take its fixpoint.

The reified approach allows him to define a grammar's semantics in terms
of bags of words and prove sound a variant of Brzozowski derivatives~\citeyear{brzozowski1964derivatives}
as well as study the equational theory of parsers. The current
implementation, based on the Brzozowski derivatives, is however of complexity
at least exponential in the size of the input.

Our approach, although not as successful as Danielsson's when it comes
to certification, is however more lightweight. Using only strong induction
on the natural numbers, it is compatible with languages a lot less powerful
than Agda. Indeed there is no need for good support for mixed induction and
coinduction in the host language. And although we do rely on enforcing invariants
at the type-level, one could mimic these in languages with even weaker type
systems by defining an \emph{abstract} \BOX{} and only providing the user with
our set of combinators which is guaranteed to be safe.

\subsection{Certified Parsing}

Ambitious projects such as CompCert~\cite{leroy2012compcert} providing
the user with an ever more certified toolchain tend to bring to light
the lack of proven-correct options for very practical concerns such as
parsing. Jourdan, Pottier and Leroy's work~\cite{jourdan2012validating}
fills that gap by certifying the output of an untrusted parser generator
for LR(1) grammars. This approach serves a different purpose than ours:
parser combinators libraries are great for rapid prototyping and small,
re-configurable parsers for non-critical applications.

Bernardy and Jansson have implemented in Agda a fully-certified generalisation
of Valiant's algorithm~\cite{bernardy2016certified} by deriving it from
its specification. This algorithm gives the best asymptotic bounds on
context-free grammar, that is the \type{Applicative} subset tackled by
parser combinators.

\section{Conclusion and Future Work}

Starting from the definition of ``parsers for things as functions from
strings to lists of strings and things'' common in Haskell, we have been
able to (re)define versatile combinators. However the type system was
completely unable to root out some badly-behaved programs, namely the
ones taking the fixpoint of a grammar accepting the empty word or non
well-founded left recursive grammars. Wanting to use a total programming
language, this led us to a radical solution: rejecting all the parsers
accepting the empty word. Luckily, it was still possible to recover a
notion of ``potentially failing'' sub-parses via a \textit{bind}-like
combinator as well as defining combinators for left chains. Finally we
saw that this yielded a perfectly safe and only barely more verbose set
of total parser combinators.

In the process of describing our library we have introduced a set of
type-level combinators for manipulating indexed types and defining
values by strong induction. If we want to provide our users with the
tools to modularly prove some of the properties of their grammars, we
need to come up with proof combinators corresponding to the value ones.
As far as we know this is still an open problem.
